---
title: "Creating a list of studied specimens"
output:
  html_notebook:
    df_print: paged
---

In this notebook, we will load a table with information on specimens studied and format it as rich text to include in the manuscript.

Let's start by loading packages:
```{r}
rm(list=ls())
library(tidyverse)
library(rtf)
library(readxl)
```

Now, let's read the dataset

```{r}
specimens = readxl::read_excel('especimes estudados_2020_rev1.xlsx')
specimens
```

First, let's consolidate the field of collector
```{r}
specimens = specimens %>%
  mutate(collector1 = ifelse(!is.na(collector_interpreted),
                            collector_interpreted,
                            colletor)
         )

specimens
```

Now, let's consolidate the collection. Some specimens were donated to other collections during the work, and we will record the destination collection.
```{r}
specimens = specimens %>%
  mutate(collection1 = ifelse(!is.na(collection_new),
                            collection_new,
                            collection)
         )

specimens

```

Now let's change sex by f# and m# that we will replace later. If this is for a paper in Zootaxa, this is the final format. Unfortunately, word can't read UTF8 encoding properly, and rtf package can't deal with other encodings, so we cannot use ♀ and ♂ directly. One has to find and replace later in the document if these are the desired symbols
```{r}
specimens = specimens %>%
  mutate(sex = ifelse(sex == 'm','m#',sex),
         sex = ifelse(sex == 'f','f#',sex),
         sex = ifelse(is.na(sex),'sex unknown',sex))

specimens
```

Now let's reformat country, state and locality.

```{r}
specimens = specimens %>%
  mutate(country = ifelse(country=='no data', 'No country', country),
         state = ifelse(state=='no data', 'No region', state),
         locality = ifelse(locality=='no data', 'No locality', locality),
         locality = ifelse(!is.na(updated_locality),str_c(locality,
                                                          ' [',
                                                          updated_locality,
                                                          ']'),
                           locality),
         collector1 = ifelse(collector1=='no data', 'no collector', collector1))
```


Now let's summarise dates. For that, let's create a function that will summarize one line.
```{r}
get_summarized_date = function(line){
  get_single_date = function(d, m, y){
    if (!is.na(d)){
        #day/month/year, single date
        
        return(str_c(sprintf("%02d",d), 
                                    as.roman(m),
                                    y,
                                    sep='/'))
      } else {
        if (!is.na(m)){
          #month/year, single date
          
          return(str_c(as.roman(m),
                                    y,
                                    sep='/'))
        } else {
          if (!is.na(y)){
            #year, single date
            return(as.character(y))
          } else {
            return('no date')
          }
          
        }
      }
  }
  
  
  if (line$is_date_legible == 'F'){
    return('date illegible')
    
  } else {
    if (is.na(line$year_end)){
      return(get_single_date(line$day_start,
                             line$month_start,
                             line$year_start))
      
    } else {
      first = get_single_date(line$day_start,
                             line$month_start,
                             line$year_start)
      second = get_single_date(line$day_end,
                             line$month_end,
                             line$year_end)
      
      y = line$year_end
      m = as.roman(line$month_end)
      first = str_remove(first,str_c('/',y))
      first = str_remove(first,str_c('/',m))
      
      return(str_c(first,
                   second,
                   sep='\U2013' #ndash
                   )
             )
    }
  }
}
```


Now we apply the function.
```{r}
specimens$date_final = specimens %>%
  split(1:nrow(.)) %>%
  purrr::map(get_summarized_date) %>%
  unlist

specimens
```


Now let's keep only the columns that we will use
```{r}
specimens = specimens %>%
  select(accepted_name, 
         type_name, 
         sex, 
         collector = collector1,
         country,
         state,
         locality,
         date = date_final,
         collection = collection1,
         type_status,
         quantity)

specimens
```

Now let's sum all specimens for each combination of locality / date / collector / sex / type status / collection.

```{r}
specimens = specimens %>% group_by(accepted_name,
                                   type_name,
                                   sex,collector,
                                   country,
                                   state,
                                   locality,
                                   date,
                                   collection,
                                   type_status) %>%
  summarise(quantity = sum(quantity))
specimens
```

The table includes some entries with quantity of 0, which were used to facilitate typing. Let's remove them now.
```{r}
specimens = specimens %>%
  filter(quantity > 0)
```



Now let's order the data frame.
```{r}
specimens = specimens %>%
  arrange(accepted_name,
          type_status,
          country,
          state,
          locality,
          collector,
          date,
          collection,
          sex)

specimens
```

Let's separate types and nontypes:
```{r}
types = specimens %>% filter(!is.na(type_status))
nontypes = specimens %>% filter(is.na(type_status)) %>%
  ungroup %>%
  select(-type_name, -type_status)
```


Now let's loop through species and make separate lists for types and non types

```{r}
rtf = RTF('material_examined.rtf')
for (sp in unique(specimens$accepted_name)){
  addHeader(rtf,sp)
  
  addHeader(rtf,'Type material')
  
  this = filter(types,accepted_name == sp)
  this$same = this$accepted_name == this$type_name
  this = this %>%
    arrange(desc(same), type_name, type_status, country, state, locality, date, sex, collection)
  
  addText(rtf,str_c(toupper(this[1,]$country),'. '),bold = F)
  addText(rtf,str_c(this[1,]$state,': '),bold = T)
  addText(rtf,str_c(this[1,]$locality,', '),bold = F)
  addText(rtf,str_c(this[1,]$collector,', '),bold = F)
  addText(rtf,str_c(this[1,]$date,' ('),bold = F)
  if (!(this[1,]$type_status %in% c('lectotype','holotype'))){
    addText(rtf,str_c(this[1,]$quantity,' '),bold = F)
  }
  addText(rtf,str_c(this[1,]$sex,' '),bold = F)
  addText(rtf,str_c(this[1,]$type_status,' '),bold = F)
  addText(rtf,str_c('Anchylorhynchus ',this[1,]$type_name,' '),italic = T)
  addText(rtf,str_c(this[1,]$collection),bold = F)
  
  previous = list(country = this[1,]$country,
                  state = this[1,]$state,
                  locality = this[1,]$locality,
                  collector = this[1,]$collector,
                  date = this[1,]$date)
  
  
  
  
  if (nrow(this) >= 2){
    for (i in 2:nrow(this)){
      if (this[i,]$country != previous$country){
        addText(rtf,').\n',bold=F)
        addText(rtf,str_c(toupper(this[i,]$country),'. '),bold = F)
        addText(rtf,str_c(this[i,]$state,': '),bold = T)
        addText(rtf,str_c(this[i,]$locality,', '),bold = F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        if (!(this[i,]$type_status %in% c('lectotype','holotype'))){
          addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        }
        addText(rtf,str_c(this[i,]$sex,' '),bold = F)
        addText(rtf,str_c(this[i,]$type_status,' '),bold = F)
        addText(rtf,str_c('Anchylorhynchus ',this[i,]$type_name,' '),italic = T)
        addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$state != previous$state){
        addText(rtf,'). ',bold=F)
        addText(rtf,str_c(this[i,]$state,': '),bold = T)
        addText(rtf,str_c(this[i,]$locality,', '),bold = F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        if (!(this[i,]$type_status %in% c('lectotype','holotype'))){
          addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        }
        addText(rtf,str_c(this[i,]$sex,' '),bold = F)
        addText(rtf,str_c(this[i,]$type_status,' '),bold = F)
        addText(rtf,str_c('Anchylorhynchus ',this[i,]$type_name,' '),italic = T)
        addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$locality != previous$locality){
        addText(rtf,'); ',bold=F)
        addText(rtf,str_c(this[i,]$locality,', '),bold = F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        if (!(this[i,]$type_status %in% c('lectotype','holotype'))){
          addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        }
        addText(rtf,str_c(this[i,]$sex,' '),bold = F)
        addText(rtf,str_c(this[i,]$type_status,' '),bold = F)
        addText(rtf,str_c('Anchylorhynchus ',this[i,]$type_name,' '),italic = T)
        addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$collector != previous$collector){
        addText(rtf,'); ',bold=F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        if (!(this[i,]$type_status %in% c('lectotype','holotype'))){
          addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        }
        addText(rtf,str_c(this[i,]$sex,' '),bold = F)
        addText(rtf,str_c(this[i,]$type_status,' '),bold = F)
        addText(rtf,str_c('Anchylorhynchus ',this[i,]$type_name,' '),italic = T)
        addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$date != previous$date){
        addText(rtf,'); ',bold=F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        if (!(this[i,]$type_status %in% c('lectotype','holotype'))){
          addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        }
        addText(rtf,str_c(this[i,]$sex,' '),bold = F)
        addText(rtf,str_c(this[i,]$type_status,' '),bold = F)
        addText(rtf,str_c('Anchylorhynchus ',this[i,]$type_name,' '),italic = T)
        addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else {
        addText(rtf,', ',bold=F)
        if (!(this[i,]$type_status %in% c('lectotype','holotype'))){
          addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        }
        addText(rtf,str_c(this[i,]$sex,' '),bold = F)
        addText(rtf,str_c(this[i,]$type_status,' '),bold = F)
        addText(rtf,str_c('Anchylorhynchus ',this[i,]$type_name,' '),italic = T)
        addText(rtf,str_c(this[i,]$collection),bold = F)
      }
      previous = list(country = this[i,]$country,
                      state = this[i,]$state,
                      locality = this[i,]$locality,
                      collector = this[i,]$collector,
                      date = this[i,]$date)
      
    }
  }
  
  addText(rtf,')\n\n',bold = F)
  
  addHeader(rtf,'Other material')
  
  this = filter(nontypes,accepted_name == sp)
  
  addText(rtf,str_c(toupper(this[1,]$country),'. '),bold = F)
  addText(rtf,str_c(this[1,]$state,': '),bold = T)
  addText(rtf,str_c(this[1,]$locality,', '),bold = F)
  addText(rtf,str_c(this[1,]$collector,', '),bold = F)
  addText(rtf,str_c(this[1,]$date,' ('),bold = F)
  addText(rtf,str_c(this[1,]$quantity,' '),bold = F)
  addText(rtf,this[1,]$sex,bold = F)
  
  
  previous = list(country = this[1,]$country,
                  state = this[1,]$state,
                  locality = this[1,]$locality,
                  collector = this[1,]$collector,
                  date = this[1,]$date,
                  collection = this[1,]$collection)
  
  
  
  
  if (nrow(this) >= 2){
    for (i in 2:nrow(this)){
      if (this[i,]$country != previous$country){
        addText(rtf,str_c(' ',previous$collection),bold = F)
        addText(rtf,').\n',bold=F)
        addText(rtf,str_c(toupper(this[i,]$country),'. '),bold = F)
        addText(rtf,str_c(this[i,]$state,': '),bold = T)
        addText(rtf,str_c(this[i,]$locality,', '),bold = F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        #addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$state != previous$state){
        addText(rtf,str_c(' ',previous$collection),bold = F)
        addText(rtf,'). ',bold=F)
        addText(rtf,str_c(this[i,]$state,': '),bold = T)
        addText(rtf,str_c(this[i,]$locality,', '),bold = F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        #addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$locality != previous$locality){
        addText(rtf,str_c(' ',previous$collection),bold = F)
        addText(rtf,'); ',bold=F)
        addText(rtf,str_c(this[i,]$locality,', '),bold = F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        #addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$collector != previous$collector){
        addText(rtf,str_c(' ',previous$collection),bold = F)
        addText(rtf,'); ',bold=F)
        addText(rtf,str_c(this[i,]$collector,', '),bold = F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        #addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$date != previous$date){
        addText(rtf,str_c(' ',previous$collection),bold = F)
        addText(rtf,'); ',bold=F)
        addText(rtf,str_c(this[i,]$date,' ('),bold = F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        #addText(rtf,str_c(this[i,]$collection),bold = F)
        
      } else if (this[i,]$collection != previous$collection) {
        addText(rtf,str_c(' ',previous$collection),bold = F)
        addText(rtf,', ',bold=F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        
      } else {
        addText(rtf,', ',bold=F)
        addText(rtf,str_c(this[i,]$quantity,' '),bold = F)
        addText(rtf,this[i,]$sex,bold = F)
        #addText(rtf,str_c(this[i,]$collection),bold = F)
      }
      previous = list(country = this[i,]$country,
                      state = this[i,]$state,
                      locality = this[i,]$locality,
                      collector = this[i,]$collector,
                      date = this[i,]$date,
                      collection = this[i,]$collection)
      
    }
  }
  addText(rtf,str_c(' ',previous$collection),bold = F)
  addText(rtf,')',bold = F)
  
  addPageBreak(rtf)
}

```

```{r}
done(rtf)
```


# To do manually:

Since measurement table and specimen table are distinct, we will add "dissected" manually.

We will also use BBedit to change encoding from UTF8 to Mac OS Roman before opening in Word.

Finally, double check the species list for anything odd.
